plyranges: a grammar for transforming genomics data

### 2018-04-13 paper.Rmd
reordering design section

The most general Bioconductor data structure, GRanges, is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, stand and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment.  For example, a gene can be represented by its chromosomal
coordinates, along with its identifier and the identifiers of its
exons; or an RNA-seq experiment may be represented as sets of genes
with a matching count column.

GRanges follows the tidy data pattern: it is a rectangular table
corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that
observation.

via
 This interface
provides abstractions for representing genomics data and enables interoperability
by allowing users and developers to use these abstractions in their own code
and packages. However, due to the large number of abstractions there are costs
involved. Due to the OO interface each abstraction will behave slightly 
differently for any given generic method. There is a cognitive load on the user
to understand the appropriate context for each abstraction. An alternative approach
is to recognise that most genomic data analysis tasks can be performed with 
data structures that naturally represent genomic intervals. That is, genomics
data can be represented as sets of integer pairs, where each pair has a start
and an end coordianate, along with strings representing strand and chromosome 
location. Measurements or metadata can be included as additional annotations
on each pair.  As an example. a gene can be represented with its coordinates, along with its 
identifier and the identifiers of its exons; or an RNA-seq experiment may be
represented as sets of genes with a matching count column.

The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
The package defines a consistent syntax and style for transforming tabular data sets via
functions that are named as verbs. Each function performs an action on the data
set and is cohesive and endomorphic. Cohesion means that each function performs
a singular task and has no side-effects, while endomorphism means a function
call returns the same type of object as its input. Taken together these
features enable a user to write human readable analysis workflows.

The `plyranges` DSL is built on two fundamental data structures defined
by the `IRanges` and `GenomicRanges` packages called _IntegerRanges_
and _GenomicRanges_. Unless otherwise specified, throughout the paper
we refer to either of these data structures as _Ranges_. In its simplest form 
the data model for an _IntegerRanges_ object is defined as a table with
$n$ rows, where each row contains a closed interval of two integers:

$$
\textit{IntegerRange} := \{ \left[\text{start}_i, \text{end}_{i}\right]  : i = 1\dots n \}.
$$
In this form an _IntegerRanges_ object has three core columns: a start, an
end, and the width of the interval. The _GenomicRanges_ data
model has additional semantics providing biological context, strand and 
sequence name (seqname), as well as containing an _IntegerRanges_ object.

$$
\textit{GenomicRange} := \{  \{\text{seqname}_i, \left[\text{start}_i, \text{end}_{i}\right] , \text{strand}_i\}  : i = 1\dots n \}.
$$
A _GenomicRanges_ object can have an additional property that provides
annotation for each sequence name over a genome. This contstrains the the 
start and end points for a given sequence to lie inside the bounds
of the genome space. Also note that _GenomicRanges_ have one-based 
coordinates rather than zero-based coordinates.  Additional measurements
or annotations can be bound to the bare _IntegerRanges_ or _GenomicRanges_ to
form metadata columns.

By definition the _Ranges_ class encapsulate tidy data prinicples because
every row corresponds to a sequence and every column corresponds to a variable.
Hence, we have designed the `plyranges` DSL to have the grammar
and design principles of `dplyr`: fluency, consistency, cohesion and 
endomorphism. All of these principles are discussed with respect to the 
_Ranges_ data model below. Where possible we contrast our design to the existing
Bioconductor infrastructure.

@@

## Overlapping Ranges

@@

@@

As an example, we use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project @Roadmap_Epigenomics_Consortium2015-pr.
We choose to focus on assays for primary T CD8+ memory cells from peripheral 
blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
 trimethylation (H3K27Me3) 
methylation mark over chromosome 10.

@@

The `reduce_ranges()` operation is used to find coverage peaks 
across chromosome 10. We can manually set a threshold to restrict genomic
regions to have a coverage score of greater than 8, and
then merge nearby regions. The maximum coverage is computed over all the
coverage scores in regions that were reduced.

@@

Returning to the _Ranges_ object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position by anchoring and 
modifying the the width.

@@

Finally, the overlap inner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contains
the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(chr10_max_score_region %>% 
                       select(-score))

@@

```{r peak-viz, echo = FALSE, fig.cap = "Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project."}
ggbio::autoplot(peak_region, 
                aes(y = score), 
                geom = "blank") + 
  geom_area(aes(x = start))
```

>>>>

The `plyranges` DSL is built on the most general Bioconductor data structure, 
GRanges, which is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, start and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment. For example a GRanges can be used to represent a gene with
its chromosomal coordinates, and along with its exon identifiers; or 
a GRanges can represent an RNA-seq experiment as a set of genes
with a corresponding count column.

By definition the GRanges follow  the tidy data pattern: it is a rectangular 
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that
observation. Hence, we have designed the `plyranges` DSL to extend the grammar
and design principles of `dplyr`:  cohesion, consistency, endomorphism,
and fluency. All of these principles are defined and discussed
below in the context of the GRanges class. Where applicable we contrast our 
design to the existing Bioconductor infrastructure.

@@

A function is cohesive if it performs a singular task and does not produce
any side-effects. In the `plyranges` DSL our algebra for performing
genomic arithmetic is a key example of cohesion. We define anchoring operators
that decorate a GRanges object with an 'anchor' and modify the semantics
of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
object by its start, center, or end coordinates (or fixing these coordinates
by strand). This enables any arithmetic function that performs a 
coordinate transformation to remain cohesive: that is they always perfom
the same operation regardless of whether a GRanges object is anchored or not.
The only difference is the result of performing the arithimetic changes when
contextual information is added to the GRanges object.

Another example of altering object semantics, while maintaining cohesion is
through the 'group_by' operator. Like anchoring, this operator decorates
a GRanges object with a column name (or names) that imply a partitioning of the 
GRanges by the unique values in the column(s). Functions defined in the 
`plyranges` DSL that perform restriction, aggregation, or column modfication
still perform those singular tasks, however grouping changes how those tasks
are performed.

@@

## Limitations

@@

## Peak Finding

We use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
@Roadmap_Epigenomics_Consortium2015-pr. We choose to focus on assays for primary
T CD8+ memory cells from peripheral blood. We can then read the BigWig file 
corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation mark 
over chromosome 10.

@@

The `reduce_ranges()` operation is used to find coverage peaks 
across chromosome 10. We can manually set a threshold to restrict genomic
regions to have a coverage score of greater than 8, and
then merge nearby regions. The maximum coverage is computed over all the
coverage scores in the regions that were reduced.

@@

Returning to the GRanges object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing the maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position by anchoring and 
modifying the the width.

@@

Finally, the overlap inner join is used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contain
the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(chr10_max_score_region)

@@

```{r peak-viz, echo = FALSE, fig.cap = "Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project."}
ggbio::autoplot(peak_region, 
                aes(y = score.x), 
                geom = "blank") + 
  geom_area(aes(x = start))
```

## Computing Windowed Statistics

(example of group_by_overlaps + summarise)

## Transition Transversion Ratio

(example of group_by + mutate + summarise and working on an object that inherits
from GRanges)
